---
title: "What factors are significantly associated with coronary heart disease diagnosis?"
author: "Team Confusion Matrix Mafia : Dhanush, Heng Jui, Nikhil, Adithya"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
library(ezids)
knitr::opts_chunk$set(warning = F, message = F)
options(scientific=T, digits = 3) 
```

# Abstract

This study examines key factors associated with coronary heart disease using simple statistical analysis. We investigate 5 main research questions using t-tests, chi-square tests, and ANOVA to understand the relationship between lifestyle factors, health metrics, and coronary heart disease diagnosis.

**Research Questions:**
1. Is there a significant difference in BMI between people with and without coronary heart disease? (T-test)
2. Is smoking status associated with coronary heart disease? (Chi-square test)
3. Does BMI vary significantly across different general health status levels? (ANOVA)
4. Does exercise participation vary significantly across different BMI categories, suggesting a role in weight management? (Chi-square test)
5. Is diabetes status associated with coronary heart disease? (Chi-square test)

# Data Loading
```{r}
Heart_data <- data.frame(read.csv("BRFSS_2024_Readable_Columns.csv"))
str(Heart_data)
summary(Heart_data)
```
# Confirm missing value ratio
```{r}
cat("=== MISSING VALUE ANALYSIS FOR STATISTICAL DECISION MAKING ===\n")

# Check missing percentages for all variables to make informed decisions
missing_analysis <- data.frame(
  Variable = names(Heart_data),
  Missing_Count = sapply(Heart_data, function(x) sum(is.na(x))),
  Missing_Percent = sapply(Heart_data, function(x) round((sum(is.na(x)) / length(x)) * 100, 2))
)

# Sort by missing percentage
missing_analysis <- missing_analysis[order(-missing_analysis$Missing_Percent), ]
print(missing_analysis)

# Focus on the four continuous variables
continuous_vars <- c("BMI_Category", "Exercise_Past_30_Days", "Mental_Health_Days", "Poor_Physical_Health_Days")

cat("\n=== CONTINUOUS VARIABLES MISSING VALUE RATES ===\n")
for(var in continuous_vars) {
  if(var %in% names(Heart_data)) {
    missing_count <- sum(is.na(Heart_data[[var]]))
    missing_pct <- (missing_count / nrow(Heart_data)) * 100
    cat(sprintf("• %s: %d missing (%.1f%%)\n", var, missing_count, missing_pct))
  }
}

# Categorize variables by missing percentage for appropriate handling
high_missing <- missing_analysis$Variable[missing_analysis$Missing_Percent > 50]
moderate_missing <- missing_analysis$Variable[missing_analysis$Missing_Percent >= 5 & missing_analysis$Missing_Percent <= 50]
low_missing <- missing_analysis$Variable[missing_analysis$Missing_Percent < 5]

cat("\nVariables categorized by missing percentage:\n")
cat("HIGH (>50% missing - should DROP):", paste(high_missing, collapse = ", "), "\n")
cat("MODERATE (5-50% missing - consider imputation):", paste(moderate_missing, collapse = ", "), "\n") 
cat("LOW (<5% missing - complete case analysis):", paste(low_missing, collapse = ", "), "\n")
```

# Data Transformation
```{r}
# Transform all variables that exist in the dataset for potential analysis

# Age groups
Heart_data$Age_Group_5yr <- factor(
  Heart_data$Age_Group_5yr,
  levels = 1:13,
  labels = c(
    "18-24", "25-29", "30-34", "35-39", "40-44", 
    "45-49", "50-54", "55-59", "60-64", "65-69", 
    "70-74", "75-79", "80+")
)

# Demographics
Heart_data$Sex <- factor(
  Heart_data$Sex,
  levels = c(1, 2),
  labels = c("Male", "Female")
)

# Health conditions
Heart_data$Diabetes_Status <- factor(
  Heart_data$Diabetes_Status,
  levels = c(1, 2, 3, 4),
  labels = c("Yes", "Yes/borderline", "Yes/pregnancy only", "No")
)

# Smoking variables
Heart_data$Smoking_Status <- factor(
  Heart_data$Smoked_100_Cigarettes_Lifetime,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

# BMI_Category_Alt 
Heart_data$BMI_Category_Alt  <- factor(
  Heart_data$BMI_Category_Alt ,
  levels = c(1, 2, 3, 4),
  labels = c("Underweight", "Normal Weight", "Overweight", "Obese")
)

# Exercise variables
Heart_data$Exercise_Past_30_Days <- factor(
  Heart_data$Exercise_Past_30_Days,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

# Cardiovascular history
Heart_data$CVD_Stroke_History <- factor(
  Heart_data$CVD_Stroke_History,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

Heart_data$Coronary_Heart_Disease <- factor(
  Heart_data$Coronary_Heart_Disease,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

Heart_data$Heart_Attack_History <- factor(
  Heart_data$Heart_Attack_History,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

# General health
Heart_data$General_Health_Status <- factor(
  Heart_data$General_Health_Status,
  levels = c(1, 2, 3, 4, 5),
  labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")
)

# BMI transformation (convert from coded value to actual BMI)
Heart_data$BMI_Value <- Heart_data$BMI_Category / 100

# Additional variables that exist in dataset
if("Flu_Vaccination" %in% names(Heart_data)) {
  Heart_data$Flu_Vaccination <- factor(
    Heart_data$Flu_Vaccination,
    levels = c(1, 2),
    labels = c("Yes", "No")
  )
}

if("Difficulty_Walking_Stairs" %in% names(Heart_data)) {
  Heart_data$Difficulty_Walking_Stairs <- factor(
    Heart_data$Difficulty_Walking_Stairs,
    levels = c(1, 2),
    labels = c("Yes", "No")
  )
}

str(Heart_data)
summary(Heart_data)
```

# Data Cleaning

## Statistically Sound Missing Value Handling

```{r}
cat("=== APPLYING EVIDENCE-BASED MISSING VALUE TREATMENT ===\n")

# Step 1: Drop variables with >50% missing (statistically unreliable)
Heart_clean <- Heart_data

# Check which variables have >50% missing and should be dropped
high_missing_vars <- c()
for(var in names(Heart_clean)) {
  missing_pct <- (sum(is.na(Heart_clean[[var]])) / nrow(Heart_clean)) * 100
  if(missing_pct > 50) {
    high_missing_vars <- c(high_missing_vars, var)
  }
}

if(length(high_missing_vars) > 0) {
  cat("Dropping variables with >50% missing:", paste(high_missing_vars, collapse = ", "), "\n")
  Heart_clean <- Heart_clean[, !names(Heart_clean) %in% high_missing_vars]
}

# Step 2: Handle continuous variables with 5-20% missing (ONLY these get imputed)
continuous_vars <- c("BMI_Value", "Mental_Health_Days", "Poor_Physical_Health_Days", "Alcohol_Days_Per_Month", "Binge_Drinking_Episodes", "Max_Drinks_Single_Occasion")
available_continuous <- continuous_vars[continuous_vars %in% names(Heart_clean)]

cat("\n=== CONTINUOUS VARIABLE IMPUTATION (5-20% missing only) ===\n")
for(var in available_continuous) {
  missing_count <- sum(is.na(Heart_clean[[var]]))
  missing_pct <- (missing_count / nrow(Heart_clean)) * 100
  
  if(missing_pct >= 5 && missing_pct <= 20) {
    # Use MEDIAN imputation (more robust than mean)
    var_median <- median(Heart_clean[[var]], na.rm = TRUE)
    Heart_clean[[var]][is.na(Heart_clean[[var]])] <- var_median
    cat(sprintf(" %s: Imputed %d values (%.1f%%) with MEDIAN (%.2f)\n", 
                var, missing_count, missing_pct, var_median))
  } else if(missing_pct > 20) {
    cat(sprintf(" %s: %.1f%% missing - Consider advanced imputation or dropping\n", var, missing_pct))
  } else {
    cat(sprintf(" %s: %.1f%% missing - Keeping as NA for complete case analysis\n", var, missing_pct))
  }
}

# Step 3: For categorical variables - NO imputation, use complete case analysis
cat("\n=== CATEGORICAL VARIABLES - NO IMPUTATION ===\n")
categorical_vars <- names(Heart_clean)[sapply(Heart_clean, function(x) is.factor(x) || is.character(x))]
for(var in categorical_vars) {
  missing_count <- sum(is.na(Heart_clean[[var]]))
  missing_pct <- (missing_count / nrow(Heart_clean)) * 100
  if(missing_count > 0) {
    cat(sprintf(" %s: %d missing (%.1f%%) - Will use complete case analysis\n", var, missing_count, missing_pct))
  }
}

# Step 4: Remove rows with missing CRITICAL variables only
critical_vars <- c("Coronary_Heart_Disease", "General_Health_Status", "Sex")
available_critical <- critical_vars[critical_vars %in% names(Heart_clean)]

# Remove invalid General Health Status responses (7=Don't know, 9=Refused) before final cleaning
if("General_Health_Status" %in% names(Heart_clean)) {
  valid_health_responses <- Heart_clean$General_Health_Status %in% c("Excellent", "Very Good", "Good", "Fair", "Poor")
  invalid_health_count <- sum(!valid_health_responses, na.rm = TRUE)
  if(invalid_health_count > 0) {
    cat(sprintf("Removing %d rows with invalid General Health Status responses\n", invalid_health_count))
    Heart_clean <- Heart_clean[valid_health_responses, ]
  }
}

# Step 5: Check rows with missing BMI_Category_Alt but having BMI_Value (Used in one research question)
if("BMI_Category_Alt" %in% names(Heart_clean)) {
  missing_bmi_cat <- sum(is.na(Heart_clean$BMI_Category_Alt) & !is.na(Heart_clean$BMI_Value))
  if(missing_bmi_cat > 0) {
    cat(sprintf("Note: %d rows missing BMI_Category_Alt but having BMI_Value - will be excluded in relevant analyses\n", missing_bmi_cat))
  }

  # Impute missing BMI_Category_Alt based on BMI_Value (standard cutoffs)
  if("BMI_Category_Alt" %in% names(Heart_clean) && "BMI_Value" %in% names(Heart_clean)) {
    missing_idx <- which(is.na(Heart_clean$BMI_Category_Alt) & !is.na(Heart_clean$BMI_Value))
    if(length(missing_idx) > 0) {
      bmi_vals <- Heart_clean$BMI_Value[missing_idx]
      # Standard BMI cutoffs
      bmi_cat <- cut(bmi_vals,
                    breaks = c(-Inf, 18.5, 24.9, 29.9, Inf),
                    labels = c("Underweight", "Normal Weight", "Overweight", "Obese"),
                    right = TRUE)
      Heart_clean$BMI_Category_Alt[missing_idx] <- bmi_cat
      cat(sprintf("Filled %d missing BMI_Category_Alt values based on BMI_Value.\n", length(missing_idx)))
    }
  }
}

rows_before <- nrow(Heart_clean)
Heart_clean <- Heart_clean[complete.cases(Heart_clean[available_critical]), ]
rows_after <- nrow(Heart_clean)

cat(sprintf("Rows removed due to missing CRITICAL variables: %d\n", rows_before - rows_after))
cat(sprintf("Final dataset: %d rows × %d columns\n", nrow(Heart_clean), ncol(Heart_clean)))
cat(sprintf("Data retention rate: %.1f%%\n", (rows_after / nrow(Heart_data)) * 100))

# Check remaining missing values
remaining_missing <- sapply(Heart_clean, function(x) sum(is.na(x)))
remaining_missing <- remaining_missing[remaining_missing > 0]
if(length(remaining_missing) > 0) {
  cat("\nRemaining missing values (for complete case analysis in specific tests):\n")
  print(remaining_missing)
} else {
  cat("\n No remaining missing values\n")
}

str(Heart_clean)
summary(Heart_clean)
```

## Statistically Appropriate Outlier Treatment (Removing impossible data / Dirty data for average human being)

```{r}
cat("=== OUTLIER ANALYSIS WITH CLINICAL RELEVANCE ===\n")

# Store data before outlier removal for comparison
Heart_before_outlier <- Heart_clean

# Only remove IMPOSSIBLE values, not clinical outliers
if("BMI_Value" %in% names(Heart_clean)) {
  # BMI: Only remove impossible values (<10 or >70 kg/m²)
  bmi_outliers <- which(Heart_clean$BMI_Value < 10 | Heart_clean$BMI_Value > 70)
  if(length(bmi_outliers) > 0) {
    cat(sprintf("Removing %d impossible BMI values (<10 or >70)\n", length(bmi_outliers)))
    Heart_clean <- Heart_clean[-bmi_outliers, ]
  } else {
    cat("All BMI values within plausible human range\n")
  }
}

# Health days: Should be 0-30, but we already handled this in data transformation
# Not removing high values as they may be clinically relevant

# Final data check - remove rows missing ONLY the variables we actually use in our 5 SMART research questions

analysis_vars <- c("BMI_Value", "BMI_Category_Alt", "General_Health_Status", "Coronary_Heart_Disease", 
                   "Sex", "Exercise_Past_30_Days", "Diabetes_Status", "Smoking_Status")
available_analysis_vars <- analysis_vars[analysis_vars %in% names(Heart_clean)]

rows_before_final <- nrow(Heart_clean)
Heart_clean <- Heart_clean[complete.cases(Heart_clean[available_analysis_vars]), ]
rows_after_final <- nrow(Heart_clean)

if(rows_before_final != rows_after_final) {
  cat(sprintf("Removed %d rows missing critical analysis variables\n", rows_before_final - rows_after_final))
} else {
  cat("No additional rows removed - all critical variables complete\n")
}

cat(sprintf("\nFinal cleaned dataset: %d rows × %d columns\n", nrow(Heart_clean), ncol(Heart_clean)))
cat(sprintf("Overall data retention: %.1f%% of original data\n", (nrow(Heart_clean) / nrow(Heart_data)) * 100))

# Validate our approach preserved key relationships by checking if the key variable BMI & Coronary_Heart_Disease are present in the final Dataset
if("BMI_Value" %in% names(Heart_clean) && "Coronary_Heart_Disease" %in% names(Heart_clean)) {
  bmi_by_coronary <- aggregate(BMI_Value ~ Coronary_Heart_Disease, Heart_clean, function(x) c(mean=mean(x), sd=sd(x), n=length(x)))
  cat("\nValidation - BMI by Coronary Heart Disease Status (preserved for analysis):\n")
  print(bmi_by_coronary)
}

str(Heart_clean)
```

# Descriptive Statistics

## Summary
```{r}
summary(Heart_clean)
```

## Structure
```{r}
str(Heart_clean)
```

## Data Visualizations
```{r}
library(ggplot2)
#Coronary_Heart_Disease
ggplot(Heart_clean, aes(x = Coronary_Heart_Disease)) +
  geom_bar(fill = "navy") +
  labs(title = "Distribution of Coronary Heart Disease",
       x = "Coronary Heart Disease", y = "Count")
```

From the chart of Coronary Heart Disease, we can see that most of the samples do not have this condition.


```{r}
# Diabetes Status
ggplot(Heart_clean, aes(x = Diabetes_Status)) +
  geom_bar(fill = "forestgreen") +
  labs(title = "Distribution of Diabetes Status",
       x = "Diabetes Status", y = "Count")
```

From the chart of Diabetes Status, we can see that most of the samples have some form of diabetes, especially those related to pregnancy.


```{r}
# Smoking Status
ggplot(Heart_clean, aes(x = Smoking_Status)) +
  geom_bar(fill = "red") +
  labs(title = "Distribution of Smoking Status",
       x = "Smoking Status", y = "Count")
```

From the chart of Smoking Status, we can see that the number of non-smokers is slightly higher than that of smokers.


```{r}
#General_Health_Status
ggplot(Heart_clean, aes(x = General_Health_Status)) +
  geom_bar(fill = "blue") +
  labs(title = "Distribution of General Health Status",
       x = "General Health Status", y = "Count")
```

From the chart of General Health Status, we can see that most of the samples are in good health, while the proportion of those with poor health is the smallest.


```{r}
#Exercise_Past_30_Days 
ggplot(Heart_clean, aes(x = Exercise_Past_30_Days)) +
  geom_bar(fill = "purple") +
  labs(title = "Distribution Exercise Past 30 Days",
       x = "Exercise Past 30 Days", y = "Count")
```

From the chart of Exercise in the Past 30 Days, we can see that most of the samples have exercised within the past 30 days.


```{r}
#BMI_Category
ggplot(Heart_clean, aes(x = BMI_Category_Alt)) +
  geom_bar(fill = "orange") +
  labs(title = "Distribution BMI_Category",
       x = "BMI Category", y = "Count")


```

From the chart of BMI Category (Alt), we can see that most of the samples fall within the normal and obese BMI ranges.


```{r}
# BMI_Value
ggplot(na.omit(Heart_clean), aes(x = BMI_Value)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(title = "Distribution of BMI",
       x = "BMI_Value", y = "Count")
```

From the chart of BMI Value, we can see that the distribution is right-skewed, with most of the data concentrated in the 20–40 range.


# Data Analysis

## Question 1: Is there a significant difference in BMI between people with and without coronary heart disease?

**Statistical Test:** T-test (comparing means of two groups)

```{r}
# Basic descriptive statistics
aggregate(BMI_Value ~ Coronary_Heart_Disease, data = Heart_clean, 
          FUN = function(x) c(mean = mean(x, na.rm = TRUE), 
                              sd = sd(x, na.rm = TRUE),
                              count = length(x[!is.na(x)])))

# Box plot visualization
boxplot(BMI_Value ~ Coronary_Heart_Disease, data = Heart_clean,
        main = "BMI by Coronary Heart Disease Status",
        xlab = "Coronary Heart Disease", ylab = "BMI",
        col = c("lightblue", "coral"))

# T-test
t_test_bmi <- t.test(BMI_Value ~ Coronary_Heart_Disease, data = Heart_clean)
print(t_test_bmi)

# Simple interpretation
if(t_test_bmi$p.value < 0.05) {
  cat("Result: There IS a significant difference in BMI between groups (p < 0.05)")
} else {
  cat("Result: There is NO significant difference in BMI between groups (p >= 0.05)")
}
```

For Q1, the t-test shows a very strong difference in BMI between people with and without coronary heart disease (p < 2e-16). This means it is almost impossible that the result happened by chance, suggesting that individuals with coronary heart disease tend to have higher BMI on average.

Although the boxplot shows similar BMI distributions between the two groups, the t-test result (p < 0.05) indicates a significant difference. This may be due to the large sample size detecting small mean differences.


## Question 2: Is smoking status associated with coronary heart disease?

**Statistical Test:** Chi-square test (testing association between two categorical variables)

```{r}
# Create contingency table
smoking_table <- table(Heart_clean$Smoking_Status, Heart_clean$Coronary_Heart_Disease)
print("Contingency Table:")
print(smoking_table)

# Calculate percentages
prop_table <- prop.table(smoking_table, margin = 1) * 100
print("Percentages by smoking status:")
print(round(prop_table, 1))

# Bar plot visualization
barplot(smoking_table, beside = TRUE, 
        main = "Smoking Status and Coronary Heart Disease",
        xlab = "Coronary Heart Disease", ylab = "Count",
        col = c("lightgreen", "pink"),
        legend = TRUE)

# Chi-square test
chi_test_smoking <- chisq.test(smoking_table)
print(chi_test_smoking)

# Simple interpretation
if(chi_test_smoking$p.value < 0.05) {
  cat("Result: There IS a significant association between smoking and coronary heart disease (p < 0.05)")
} else {
  cat("Result: There is NO significant association between smoking and coronary heart disease (p >= 0.05)")
}
```


For Q2, the chi-squared test shows a very strong relationship between smoking status and coronary heart disease (p < 2e-16). This means it is almost impossible that the result happened by chance, suggesting that smokers are much more likely to have coronary heart disease than non-smokers.

We can clearly see from the chart that people who smoke have a higher proportion of coronary heart disease compared to those who do not smoke.


## Question 3: Does BMI vary significantly across different general health status levels?

**Statistical Test:** ANOVA (comparing means across multiple groups)

```{r}
# Basic descriptive statistics by general health status
aggregate(BMI_Value ~ General_Health_Status, data = Heart_clean, 
          FUN = function(x) c(mean = mean(x, na.rm = TRUE), 
                              sd = sd(x, na.rm = TRUE),
                              count = length(x[!is.na(x)])))

# Box plot visualization
boxplot(BMI_Value ~ General_Health_Status, data = Heart_clean,
        main = "BMI by General Health Status",
        xlab = "General Health Status", ylab = "BMI",
        las = 2, # rotate x-axis labels
        col = c("darkgreen", "lightgreen", "yellow", "orange", "red"))

# ANOVA test
anova_bmi_health <- aov(BMI_Value ~ General_Health_Status, data = Heart_clean)
anova_summary <- summary(anova_bmi_health)
print(anova_summary)

# Simple interpretation
p_value_anova <- anova_summary[[1]][["Pr(>F)"]][1]
if(p_value_anova < 0.05) {
  cat("Result: BMI DOES vary significantly across general health status levels (p < 0.05)")
} else {
  cat("Result: BMI does NOT vary significantly across general health status levels (p >= 0.05)")
}
```

For Q3, the ANOVA test shows a very strong difference in BMI across different general health status levels (p < 2e-16). This means it is almost impossible that the result happened by chance, suggesting that BMI varies significantly depending on an individual’s reported health condition.

From the chart, we can also observe that people with poorer health tend to have higher BMI values, while those with better health tend to have lower BMI values.


## Question 4: Exercise Participation Across BMI Categories

**Research Question**: Does exercise participation vary significantly across different BMI categories, suggesting a role in weight management?

**Method**: Chi-square test of independence

```{r question4}

## Use existing BMI_Category_Alt variable for analysis, exclude NA values
bmi_exercise_table <- table(Heart_clean$BMI_Category_Alt, Heart_clean$Exercise_Past_30_Days, useNA = "no")
print("Contingency Table: BMI Categories vs Exercise Participation")
print(bmi_exercise_table)

# Calculate percentages within each BMI category
print("Exercise participation rates by BMI category:")
exercise_rates_by_bmi <- prop.table(bmi_exercise_table, margin=1) * 100
print(round(exercise_rates_by_bmi, 2))

# Perform chi-square test
chi_test_bmi_exercise <- chisq.test(bmi_exercise_table)
print(chi_test_bmi_exercise)

# Create a grouped bar plot
library(ggplot2)
bmi_exercise_df <- as.data.frame(bmi_exercise_table)
colnames(bmi_exercise_df) <- c("BMI_Category", "Exercise", "Count")

ggplot(bmi_exercise_df, aes(x = BMI_Category, y = Count, fill = Exercise)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Exercise Participation by BMI Category",
       subtitle = "Exploring the Exercise-Weight Management Connection",
       x = "BMI Category", 
       y = "Count",
       fill = "Exercise Past 30 Days") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


For Q4, the chi-squared test shows a very strong relationship between past-30-day exercise participation and BMI category (p < 2e-16). This means it is almost impossible that the result happened by chance. People who reported exercising in the past 30 days tend to fall into lower BMI categories, while those who did not exercise are more likely to be in higher BMI groups. This pattern suggests that regular physical activity plays an important role in weight management and maintaining a healthier BMI.

From the chart, we can also observe that people who are underweight or obese have a noticeably lower rate of exercise in the past 30 days compared to those with a moderate BMI.


## Question 5: Is diabetes status associated with coronary heart disease?

**Statistical Test:** Chi-square test (testing association between two categorical variables)

```{r}
# Create contingency table
diabetes_table <- table(Heart_clean$Diabetes_Status, Heart_clean$Coronary_Heart_Disease)
print("Contingency Table:")
print(diabetes_table)

# Calculate percentages
prop_table_diabetes <- prop.table(diabetes_table, margin = 1) * 100
print("Percentages by diabetes status:")
print(round(prop_table_diabetes, 1))

# Bar plot visualization
barplot(diabetes_table, beside = TRUE, 
        main = "Diabetes Status and Coronary Heart Disease",
        xlab = "Coronary Heart Disease", ylab = "Count",
        col = c("yellow", "orange", "red", "green"),
        legend = TRUE)

# Chi-square test
chi_test_diabetes <- chisq.test(diabetes_table)
print(chi_test_diabetes)

# Simple interpretation
if(chi_test_diabetes$p.value < 0.05) {
  cat("Result: There IS a significant association between diabetes and coronary heart disease (p < 0.05)")
} else {
  cat("Result: There is NO significant association between diabetes and coronary heart disease (p >= 0.05)")
}
```

For Q5, the chi-squared test shows a very strong relationship between diabetes status and coronary heart disease (p < 2e-16). This means it is almost impossible that the result happened by chance. People with diabetes or prediabetes have a much higher likelihood of developing coronary heart disease compared to those without diabetes, suggesting that poor blood sugar control is closely linked to increased heart disease risk.

Moreover, We can see from the chart that individuals with diabetes have a noticeably higher proportion of coronary heart disease compared to others.


# Conclusion

This analysis explored key lifestyle and health factors associated with coronary disease diagnosis. Statistical tests revealed strong and consistent patterns across multiple dimensions. Individuals with higher BMI, diabetes, or smoking habits showed significantly greater likelihood of having CHD. ANOVA results confirmed that BMI varies notably with self-reported health status, indicating that poor health perception aligns with higher BMI. Moreover, those who engaged in regular exercise within the past 30 days tended to maintain lower BMI categories, reinforcing the importance of physical activity in weight management and heart health.

Overall, the findings suggest that coronary heart disease is strongly linked to modifiable risk factors such as body weight, smoking, diabetes, and exercise habits. These results highlight the crucial role of healthy lifestyle choices in preventing and managing heart disease.

