---
title: "What factors are significantly associated with coronary heart disease diagnosis?"
author: " "
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
library(ezids)
knitr::opts_chunk$set(warning = F, message = F)
options(scientific=T, digits = 3) 
```

# Abstract

This study examines key factors associated with coronary heart disease using simple statistical analysis. We investigate 5 main research questions using t-tests, chi-square tests, and ANOVA to understand the relationship between lifestyle factors, health metrics, and coronary heart disease diagnosis.

**Research Questions:**
1. Is there a significant difference in BMI between people with and without coronary heart disease? (T-test)
2. Is smoking status associated with coronary heart disease? (Chi-square test)
3. Does BMI vary significantly across different general health status levels? (ANOVA)
4. Does exercise participation vary significantly across different BMI categories, suggesting a role in weight management? (Chi-square test)
5. Is diabetes status associated with coronary heart disease? (Chi-square test)

# Data Loading
```{r}
Heart_data <- data.frame(read.csv("BRFSS_2024_Readable_Columns.csv"))
str(Heart_data)
summary(Heart_data)
```
# Confirm missing value ratio
```{r}
cat("=== MISSING VALUE ANALYSIS FOR STATISTICAL DECISION MAKING ===\n")

# Check missing percentages for all variables to make informed decisions
missing_analysis <- data.frame(
  Variable = names(Heart_data),
  Missing_Count = sapply(Heart_data, function(x) sum(is.na(x))),
  Missing_Percent = sapply(Heart_data, function(x) round((sum(is.na(x)) / length(x)) * 100, 2))
)

# Sort by missing percentage
missing_analysis <- missing_analysis[order(-missing_analysis$Missing_Percent), ]
print(missing_analysis)

# Focus on the four continuous variables
continuous_vars <- c("BMI_Category", "Exercise_Past_30_Days", "Mental_Health_Days", "Poor_Physical_Health_Days")

cat("\n=== CONTINUOUS VARIABLES MISSING VALUE RATES ===\n")
for(var in continuous_vars) {
  if(var %in% names(Heart_data)) {
    missing_count <- sum(is.na(Heart_data[[var]]))
    missing_pct <- (missing_count / nrow(Heart_data)) * 100
    cat(sprintf("• %s: %d missing (%.1f%%)\n", var, missing_count, missing_pct))
  }
}

# Categorize variables by missing percentage for appropriate handling
high_missing <- missing_analysis$Variable[missing_analysis$Missing_Percent > 50]
moderate_missing <- missing_analysis$Variable[missing_analysis$Missing_Percent >= 5 & missing_analysis$Missing_Percent <= 50]
low_missing <- missing_analysis$Variable[missing_analysis$Missing_Percent < 5]

cat("\nVariables categorized by missing percentage:\n")
cat("HIGH (>50% missing - should DROP):", paste(high_missing, collapse = ", "), "\n")
cat("MODERATE (5-50% missing - consider imputation):", paste(moderate_missing, collapse = ", "), "\n") 
cat("LOW (<5% missing - complete case analysis):", paste(low_missing, collapse = ", "), "\n")
```

# Data Transformation
```{r}
# Transform all variables that exist in the dataset for potential analysis

# Age groups
Heart_data$Age_Group_5yr <- factor(
  Heart_data$Age_Group_5yr,
  levels = 1:13,
  labels = c(
    "18-24", "25-29", "30-34", "35-39", "40-44", 
    "45-49", "50-54", "55-59", "60-64", "65-69", 
    "70-74", "75-79", "80+")
)

# Demographics
Heart_data$Sex <- factor(
  Heart_data$Sex,
  levels = c(1, 2),
  labels = c("Male", "Female")
)

# Health conditions
Heart_data$Diabetes_Status <- factor(
  Heart_data$Diabetes_Status,
  levels = c(1, 2, 3, 4),
  labels = c("Yes", "Yes/borderline", "Yes/pregnancy only", "No")
)

# Smoking variables
Heart_data$Smoking_Status <- factor(
  Heart_data$Smoked_100_Cigarettes_Lifetime,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

# Exercise variables
Heart_data$Exercise_Past_30_Days <- factor(
  Heart_data$Exercise_Past_30_Days,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

# Cardiovascular history
Heart_data$CVD_Stroke_History <- factor(
  Heart_data$CVD_Stroke_History,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

Heart_data$Coronary_Heart_Disease <- factor(
  Heart_data$Coronary_Heart_Disease,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

Heart_data$Heart_Attack_History <- factor(
  Heart_data$Heart_Attack_History,
  levels = c(1, 2),
  labels = c("Yes", "No")
)

# General health
Heart_data$General_Health_Status <- factor(
  Heart_data$General_Health_Status,
  levels = c(1, 2, 3, 4, 5),
  labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")
)

# BMI transformation (convert from coded value to actual BMI)
Heart_data$BMI_Value <- Heart_data$BMI_Category / 100

# Additional variables that exist in dataset
if("Flu_Vaccination" %in% names(Heart_data)) {
  Heart_data$Flu_Vaccination <- factor(
    Heart_data$Flu_Vaccination,
    levels = c(1, 2),
    labels = c("Yes", "No")
  )
}

if("Difficulty_Walking_Stairs" %in% names(Heart_data)) {
  Heart_data$Difficulty_Walking_Stairs <- factor(
    Heart_data$Difficulty_Walking_Stairs,
    levels = c(1, 2),
    labels = c("Yes", "No")
  )
}

str(Heart_data)
summary(Heart_data)
```

# Data Cleaning

## Statistically Sound Missing Value Handling

```{r}
cat("=== APPLYING EVIDENCE-BASED MISSING VALUE TREATMENT ===\n")

# Step 1: Drop variables with >50% missing (statistically unreliable)
Heart_clean <- Heart_data

# Check which variables have >50% missing and should be dropped
high_missing_vars <- c()
for(var in names(Heart_clean)) {
  missing_pct <- (sum(is.na(Heart_clean[[var]])) / nrow(Heart_clean)) * 100
  if(missing_pct > 50) {
    high_missing_vars <- c(high_missing_vars, var)
  }
}

if(length(high_missing_vars) > 0) {
  cat("Dropping variables with >50% missing:", paste(high_missing_vars, collapse = ", "), "\n")
  Heart_clean <- Heart_clean[, !names(Heart_clean) %in% high_missing_vars]
}

# Step 2: Handle continuous variables with 5-20% missing (ONLY these get imputed)
continuous_vars <- c("BMI_Value", "Mental_Health_Days", "Poor_Physical_Health_Days", "Alcohol_Days_Per_Month", "Binge_Drinking_Episodes", "Max_Drinks_Single_Occasion")
available_continuous <- continuous_vars[continuous_vars %in% names(Heart_clean)]

cat("\n=== CONTINUOUS VARIABLE IMPUTATION (5-20% missing only) ===\n")
for(var in available_continuous) {
  missing_count <- sum(is.na(Heart_clean[[var]]))
  missing_pct <- (missing_count / nrow(Heart_clean)) * 100
  
  if(missing_pct >= 5 && missing_pct <= 20) {
    # Use MEDIAN imputation (more robust than mean)
    var_median <- median(Heart_clean[[var]], na.rm = TRUE)
    Heart_clean[[var]][is.na(Heart_clean[[var]])] <- var_median
    cat(sprintf(" %s: Imputed %d values (%.1f%%) with MEDIAN (%.2f)\n", 
                var, missing_count, missing_pct, var_median))
  } else if(missing_pct > 20) {
    cat(sprintf(" %s: %.1f%% missing - Consider advanced imputation or dropping\n", var, missing_pct))
  } else {
    cat(sprintf(" %s: %.1f%% missing - Keeping as NA for complete case analysis\n", var, missing_pct))
  }
}

# Step 3: For categorical variables - NO imputation, use complete case analysis
cat("\n=== CATEGORICAL VARIABLES - NO IMPUTATION ===\n")
categorical_vars <- names(Heart_clean)[sapply(Heart_clean, function(x) is.factor(x) || is.character(x))]
for(var in categorical_vars) {
  missing_count <- sum(is.na(Heart_clean[[var]]))
  missing_pct <- (missing_count / nrow(Heart_clean)) * 100
  if(missing_count > 0) {
    cat(sprintf(" %s: %d missing (%.1f%%) - Will use complete case analysis\n", var, missing_count, missing_pct))
  }
}

# Step 4: Remove rows with missing CRITICAL variables only
critical_vars <- c("Coronary_Heart_Disease", "General_Health_Status", "Sex")
available_critical <- critical_vars[critical_vars %in% names(Heart_clean)]

# Remove invalid General Health Status responses (7=Don't know, 9=Refused) before final cleaning
if("General_Health_Status" %in% names(Heart_clean)) {
  valid_health_responses <- Heart_clean$General_Health_Status %in% c("Excellent", "Very Good", "Good", "Fair", "Poor")
  invalid_health_count <- sum(!valid_health_responses, na.rm = TRUE)
  if(invalid_health_count > 0) {
    cat(sprintf("Removing %d rows with invalid General Health Status responses\n", invalid_health_count))
    Heart_clean <- Heart_clean[valid_health_responses, ]
  }
}

rows_before <- nrow(Heart_clean)
Heart_clean <- Heart_clean[complete.cases(Heart_clean[available_critical]), ]
rows_after <- nrow(Heart_clean)

cat(sprintf("Rows removed due to missing CRITICAL variables: %d\n", rows_before - rows_after))
cat(sprintf("Final dataset: %d rows × %d columns\n", nrow(Heart_clean), ncol(Heart_clean)))
cat(sprintf("Data retention rate: %.1f%%\n", (rows_after / nrow(Heart_data)) * 100))

# Check remaining missing values
remaining_missing <- sapply(Heart_clean, function(x) sum(is.na(x)))
remaining_missing <- remaining_missing[remaining_missing > 0]
if(length(remaining_missing) > 0) {
  cat("\nRemaining missing values (for complete case analysis in specific tests):\n")
  print(remaining_missing)
} else {
  cat("\n No remaining missing values\n")
}

str(Heart_clean)
summary(Heart_clean)
```

## Statistically Appropriate Outlier Treatment (Removing impossible data / Dirty data for average human being)

```{r}
cat("=== OUTLIER ANALYSIS WITH CLINICAL RELEVANCE ===\n")

# Store data before outlier removal for comparison
Heart_before_outlier <- Heart_clean

# Only remove IMPOSSIBLE values, not clinical outliers
if("BMI_Value" %in% names(Heart_clean)) {
  # BMI: Only remove impossible values (<10 or >70 kg/m²)
  bmi_outliers <- which(Heart_clean$BMI_Value < 10 | Heart_clean$BMI_Value > 70)
  if(length(bmi_outliers) > 0) {
    cat(sprintf("Removing %d impossible BMI values (<10 or >70)\n", length(bmi_outliers)))
    Heart_clean <- Heart_clean[-bmi_outliers, ]
  } else {
    cat("All BMI values within plausible human range\n")
  }
}

# Health days: Should be 0-30, but we already handled this in data transformation
# Not removing high values as they may be clinically relevant

# Final data check - remove rows missing ONLY the variables we actually use in our 5 SMART research questions

analysis_vars <- c("BMI_Value", "General_Health_Status", "Coronary_Heart_Disease", 
                   "Sex", "Exercise_Past_30_Days", "Diabetes_Status", "Smoking_Status")
available_analysis_vars <- analysis_vars[analysis_vars %in% names(Heart_clean)]

rows_before_final <- nrow(Heart_clean)
Heart_clean <- Heart_clean[complete.cases(Heart_clean[available_analysis_vars]), ]
rows_after_final <- nrow(Heart_clean)

if(rows_before_final != rows_after_final) {
  cat(sprintf("Removed %d rows missing critical analysis variables\n", rows_before_final - rows_after_final))
} else {
  cat("No additional rows removed - all critical variables complete\n")
}

cat(sprintf("\nFinal cleaned dataset: %d rows × %d columns\n", nrow(Heart_clean), ncol(Heart_clean)))
cat(sprintf("Overall data retention: %.1f%% of original data\n", (nrow(Heart_clean) / nrow(Heart_data)) * 100))

# Validate our approach preserved key relationships by checking if the key variable BMI & Coronary_Heart_Disease are present in the final Dataset
if("BMI_Value" %in% names(Heart_clean) && "Coronary_Heart_Disease" %in% names(Heart_clean)) {
  bmi_by_coronary <- aggregate(BMI_Value ~ Coronary_Heart_Disease, Heart_clean, function(x) c(mean=mean(x), sd=sd(x), n=length(x)))
  cat("\nValidation - BMI by Coronary Heart Disease Status (preserved for analysis):\n")
  print(bmi_by_coronary)
}

str(Heart_clean)
```

# Descriptive Statistics

## Summary
```{r}
summary(Heart_clean)
```

## Structure
```{r}
str(Heart_clean)
```

## Data Visualizations
